Flabby and injective objects in toposes

### 2022-02-14 paper-flabby-objects.tex
Full presentation of the existence of enough flabby modules

\begin{defn}A set~$X$ is \emph{flabby} if and only if every subterminal subset
of~$X$ is a subsingleton, that is, if and only if for every subset~$K \subseteq
X$ such that~$\forall x,y \in K\_ x = y$, there exists an element~$x \in X$
such that~$K \subseteq \{ x \}$.
\end{defn}

@@

\XXX{enough flabby modules}

\XXX{proper exposition}

@@

A map~$\varepsilon : \P_{\leq1}(X) \to X$ satisfies the condition in
Definition~\ref{defn:functionally-flabby} if and only if it is a retraction of
the canonical injection~$X \to \P_{\leq1}(X)$.

@@

\begin{proof}Let~$M$ be a module. First we'll assume that~$M$
is a sheaf for some modal operator~$\nabla$ such that for any
formula~$\varphi$, $\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. In
this case, the module~$M$ is already functionally flabby: A witnessing choice
function maps a subterminal~$K \subseteq M$ to the unique element~$x \in M$
such that~$\nabla(x \in K')$, where~$K' \defeq K \cup \{ 0 \,|\, \text{$K$
inhabited} \Rightarrow \nabla\bot \}$. Since~$K'$ has the property
that~$\nabla(\text{$K$ is a singleton})$, such an element~$x$ exists and is unique
by the sheaf condition.

Now let~$M$ be a module which is separated for some XXX modal operator~$\nabla$.
Then its sheafification is a functionally flabby module into which~$M$ embeds.

Finally, let~$M$ be an arbitrary module. For any element~$x \in M$,
let~$\nabla_x$ be the modal operator with~$\nabla_x\varphi \defeqv ((\varphi \Rightarrow x
= 0) \Rightarrow x = 0)$. The module~$M$ might not be~$\nabla_x$-separated for
a particular element~$x$, but jointly XXX
\end{proof}

@@

%\item There is a way of embedding any module into a flabby module if one is
%prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
%Inductively construct a collection~$T$ of terms by the following clauses: $0
%\in T$ (where~$0$ is a formal symbol);
%if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
%T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
%subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
%the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
%u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $0t \sim 0$, $1t \sim t$, $r(t+s) \sim rt
%+ rs$, $(r+r')t \sim rt + r't$ and such
%that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
%r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
%t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
%into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.

\item There is a way of embedding any module into a flabby module if
\emph{quotient inductive types}, as suggested by Altenkirch and
Kaposi~\cite{altenkirch-kaposi:qits}, are available. These generalize ordinary
inductive~$W$-types, which exist in any
topos~\cite{moerdijk-palmgren:wellfounded-trees,berg-moerdijk:w-types-in-sheaves,berg-kouwenhoven-gentil:w-types-in-eff}
and whose existence
can indeed be verified in an intuitionistic set theory like IZF, by allowing to
give constructors and state identifications at the same time. More
specifically, given an~$R$-module~$M$, we can construct a flabby envelope~$T$
of~$M$ as the quotient inductive type generated by the following clauses:~$0
\in T$ (where~$0$ is a formal symbol); if~$t,s \in T$, then~$t + s \in T$;
if~$t \in T$ and~$r \in R$, then~$rt \in T$; if~$x \in M$, then~$\underline{x}
\in T$; if~$K \subseteq T$ is a subterminal, then~$\varepsilon_K \in T$;
if~$t,s,u \in T$ and~$r,r' \in R$, then~$t + (s + u) = (t + s) + u$, $t + s = s + t$, $t + 0 = t
= t + 0$, $0t = 0$, $1t = t$, $r(t+u) = rt + ru$, $(r+r')t = rt + r't$; if~$x,y
\in M$ and~$r \in R$, then~$\underline{0} = 0$, $\underline{x + y} =
\underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$t
\in T$, then~$\varepsilon_{\{t\}} = t$.

However, it is an open question under which circumstances quotient inductive
types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
hence IZF also does not.\footnote{With quotient inductive types, every infinitary
algebraic theory admits free algebras. However, it is consistent with
Zermelo--Fraenkel set theory that some such theories do not admit free
algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
seems to be, as the existence of enough injective modules, \emph{constructively
neutral}.

\item There are a number of simple constructions which come close to providing
flabby envelopes for arbitrary modules~$M$. For instance, we could equip the set~$T
\defeq \P_{\leq1}(X)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K
\cup L \subseteq \{0\}$, with a module structure given by~$0 \defeq [\{0\}]$,
$[K]+[L] \defeq [K+L]$ and~$r [K] \defeq rK$. The resulting module admits a
linear injection from~$M$, sending an element~$x$ to~$[\{x\}]$. However, it
fails to be flabby. Given a subterminal~$E \subseteq \P_{\leq1}(X)/{\sim}$,
there is the well-defined element~$v \defeq [\{x \in M\,|\,\text{$x \in K$ for
some~$[K] \in E$}\}]$, but we cannot verify~$E \subseteq \{v\}$.

@@

>>>>

\begin{defn}\label{defn:flabby-set}A set~$X$ is \emph{flabby} if and only if every subterminal subset
of~$X$ is a subsingleton, that is, if and only if for every subset~$K \subseteq
X$ such that~$\forall x,y \in K\_ x = y$, there exists an element~$x \in X$
such that~$K \subseteq \{ x \}$.
\end{defn}

@@

\subsection{On the existence of enough flabby modules}

For the intended application to the theory of sheaf cohomology, the existence
of \emph{enough flabby sheaves of modules} is crucial: Any sheaf of modules
embeds into a flabby sheaf of modules. In this section, we study the existence
of enough flabby modules from the non-sheaf theoretic but constructive point of
view.

We have already established that every set embeds into a flabby set. However,
the supersets suggested by the proof of Corollary~\ref{cor:enough-flabby-sets}
do not carry a module structure even when the base set does. This deficit
raises the following question: Given a module~$M$, does there exist a linear
injection~$M \to M'$ into a set~$M'$ which simultaneously carries a module
structure and is flabby?

An earlier version of this text formulated this question as an open problem.
The positive solution presented below rests on the intermediate notion of
\emph{functionally flabby} sets and makes essential use of sheaves for modal
operators. A short survey for the latter is contained
in~\cite[Section~6]{blechschmidt:phd}; other references
include~\cite[Sections~14.4f.]{goldblatt:topoi},~\cite{goldblatt:modality}
\cite{vries:sheafification} and~\cite{fourman-scott:sheaves-and-logic}.\footnote{On page~5 of the
preprint~\cite{vries:sheafification} there is a slight typing error:
Fact~2.1(i) gives the characterization of~$j$-closedness, not~$j$-denseness.
The correct characterization of~$j$-denseness in that context is~$\forall b \in
B\_ j(b \in A)$.}

@@

Trivially, every functionally flabby set is flabby in the sense of
Definition~\ref{defn:flabby-set}.
A map~$\varepsilon : \P_{\leq1}(X) \to X$ satisfies the condition in
Definition~\ref{defn:functionally-flabby} if and only if it is a retraction of
the canonical injection~$X \to \P_{\leq1}(X)$.

@@

\begin{proof}We explain how to construct the required flabby envelopes
in three steps.
First, let~$M$ be a module which is a sheaf for some modal operator~$\nabla$
with the property that for any
formula~$\varphi$, $\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. In
this case, the module~$M$ is already functionally flabby: A witnessing choice
function maps a subterminal~$K \subseteq M$ to the unique element~$x \in M$
such that~$\nabla(x \in K')$, where~$K' \defeq K \cup \{ 0 \,|\, \text{$K$
inhabited} \Rightarrow \nabla\bot \}$. Since the set~$K'$ has, unlike~$K$, the property
that~$\nabla(\text{$K'$ is a singleton})$, such an element~$x$ exists and is unique
by the sheaf condition.

Second, let~$M$ be a module which is separated for some modal operator~$\nabla$
as above. Then its sheafification is a functionally flabby module into
which~$M$ embeds since the canonical map from~$M$ to its sheafification is
injective by separatedness of~$M$.

Finally, let~$M$ be an arbitrary module. For any element~$x \in M$,
let~$\nabla_x$ be the modal operator with~$\nabla_x\varphi \defeqv ((\varphi \Rightarrow x
= 0) \Rightarrow x = 0)$. This is a modal operator of the kind as above.
The module~$M$ might not be~$\nabla_x$-separated for
any particular element~$x$, but it is jointly so: For all elements~$a,b\in M$,
if~$\nabla_x(a = b)$ for all~$x \in M$, then~$a = b$ (considering~$x \defeq
a-b$). Hence the canonical map~$M \to \prod_{x \in M} M^{+_x}$ into the product
of the plus constructions with respect to all the modal operators~$\nabla_x$ is
injective, and the desired injection is the composition
\[ M \longrightarrow \prod_{x \in M} M^{+_x} \longrightarrow \prod_{x \in M}
M^{+_x+_x} \]
into the product of the sheafifications.
\end{proof}

\begin{rem}For the purpose of verifying that any module embeds into a flabby
module, it is essential that in the first step of the proof of
Theorem~\ref{thm:enough-flabby-modules} the module~$M$ is not only shown to be flabby,
but even functionally flabby, and moreover in an explicit manner, explicitly
presenting a witnessing map~$\P_{\leq1}(M) \to M$. This is because while an
arbitrary product of flabby sets can fail to be flabby, an arbitrary product of
functionally flabby sets with given witnesses is again so.
\end{rem}

Given the somewhat nontrivial nature of the construction in the proof of
Theorem~\ref{thm:enough-flabby-modules}, a natural question is whether simpler
constructions exist as well.
There are a number of simple constructions which come close to providing
flabby envelopes for arbitrary modules, but all such constructions known to the
author fail in some manner. For instance, given a module~$M$, we could equip the set~$T
\defeq \P_{\leq1}(M)/{\sim}$, where~$K \sim L$ if and only if~$K = L$ or~$K
\cup L \subseteq \{0\}$, with a module structure given by~$0 \defeq [\{0\}]$,
$[K]+[L] \defeq [K+L]$ and~$r [K] \defeq rK$. The resulting module admits a
linear injection from~$M$, sending an element~$x$ to~$[\{x\}]$. However, it
fails to be flabby. Given a subterminal~$E \subseteq \P_{\leq1}(M)/{\sim}$,
there is the well-defined element~$v \defeq [\{x \in M\,|\,\text{$x \in K$ for
some~$[K] \in E$}\}]$, but we cannot verify~$E \subseteq \{v\}$.

That said, there is an alternative construction if sufficiently general
\emph{quotient inductive types}, as suggested by Altenkirch and
Kaposi~\cite{altenkirch-kaposi:qits}, are available. These generalize ordinary
inductive~$W$-types, which exist in any
topos~\cite{moerdijk-palmgren:wellfounded-trees,berg-moerdijk:w-types-in-sheaves,berg-kouwenhoven-gentil:w-types-in-eff}
and whose existence
can indeed be verified in an intuitionistic set theory like~\textsc{izf}, by allowing to
give constructors and state identifications at the same time. More
specifically, given an~$R$-module~$M$, we can construct a flabby envelope~$T$
of~$M$ as the quotient inductive type generated by the following clauses,
starting out as the construction of the free module over the underlying set
of~$M$:~$0
\in T$ (where~$0$ is a formal symbol); if~$t,s \in T$, then~$t + s \in T$;
if~$t \in T$ and~$r \in R$, then~$rt \in T$; if~$x \in M$, then~$\underline{x}
\in T$; if~$K : I \to T$ is a family of elements of~$T$ indexed by a subterminal, then~$\varepsilon_K \in T$;
if~$t,s,u \in T$ and~$r,r' \in R$, then~$t + (s + u) = (t + s) + u$, $t + s = s + t$, $t + 0 = t
= t + 0$, $0t = 0$, $1t = t$, $r(t+u) = rt + ru$, $(r+r')t = rt + r't$; if~$x,y
\in M$ and~$r \in R$, then~$\underline{0} = 0$, $\underline{x + y} =
\underline{x} + \underline{y}$, $\underline{rx} = r \underline{x}$; and if~$K :
I \to T$ is a family such that~$I$ is inhabited by some element~$i_0$,
then~$\varepsilon_{K} = K(i_0)$.

However, there are two issues with this approach.
Firstly, it is an open question under which circumstances quotient inductive
types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
hence~\textsc{izf} also does not.\footnote{With quotient inductive types, every infinitary
algebraic theory admits free algebras. However, it is consistent with
Zermelo--Fraenkel set theory that some such theories do not admit free
algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
seems to be, as the existence of enough injective modules, \emph{constructively
neutral}.
Secondly, by referencing arbitrary families of elements, the construction
transcends the given type-theoretic or set-theoretic universe; the resulting
object~$T$ is not manifestly small even if~$M$ is.

\subsection{Exactness properties}

@@

@@

1. Zorn
2. effective
3. predicative, no AU's
